Mixture Models for Single-Cell Assays with Application to Vaccine Studies: Supplementary Materials

### 2013-05-18 MIMOSA.tex
Update revisions

During analysis, cell level biomarker fluorescence intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology.
This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$, where $N$ is the number of cells recorded and $K$ is number of biomarkers.
Using this matrix, one can form $2^K$ putative cell subsets obtained as Boolean combinations.
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty.
A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination). Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (\textit{e.g.}, animal studies).

@@

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.

>>>>

During analysis, cell level biomarker fluorescence intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology.
This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$, where $N$ is the number of cells recorded and $K$ is number of biomarkers.
Using this matrix, one can form $2^K$ putative cell subsets obtained as Boolean combinations.
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty.
A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after stimulation). Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (\textit{e.g.}, animal studies).

@@

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}. The comparisons between stimulated and unstimulated samples are typically done within a time point (\textit{i.e.,} within pre--vaccination and within post--vaccination time points), rather than directly between pre--and post--vaccine time points to avoid confounding vaccine effect with date or time of blood draw; additionally not all vaccine trials collect baseline (pre--vaccination) samples.
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.

### 2013-05-18 SupplementaryMaterialMIMOSA.tex
Update revisions

>>>>

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
    \node at (0,0) (A) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/volcanoplots.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {A} ;
	\end{scope}
     \end{tikzpicture}
     };
    \end{tikzpicture}
   \caption{Volcano plots of effect size vs posterior probability of response for each model fit to each cytokine in the ICS data set. Green points indicate subjects called responders for that cytokine and stimulation at a 1\% FDR threshold (adjusted across subjects within cytokine subset). Red points indicate non-responders.
}
   \label{webfig:simulations_Nobs}
\end{figure}
\clearpage
